---
title: "Portfolio demo with unfished reef fishes"
author: "Megsie"
date: "9/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(zoo) # for filling NAs
library(tidyverse)

# if you need it, ecofolio install:
#devtools::install_github("seananderson/ecofolio")
```

## Portfolio effects

This doc accompnaies the timesseries.Rmd presentation. I had some issues with `xaringan` so I moved the second half of the demo to a Markdown file. 

## Example with LTER fish data

### The data

Lily has prepared some LTER fish data, which look like this:
```{r echo=FALSE, eval=TRUE, tidy=FALSE}
fish <- read.csv(here::here("data","final_six_fished_unfished.csv"))
fish %>%
  mutate(fished = recode(fished, 
                         no = "unfished",
                         yes = "fished")) %>%
  ggplot(aes(x = Year, y = Count_m2, colour = Taxonomy)) +
  geom_line() +
  facet_wrap(~fished) +
  theme_classic(base_size = 16)
```

### Portfolio analysis using `ecofolio`

This requires data for every year. In the past I have dealt with this by using a state-space model to estimate states for every year, but here I just did some interpolation thing with `zoo::na.approx()` for expediency.

```{r echo=FALSE}

all <- fish %>% 
  filter(fished == "no") %>%
  expand(Taxonomy, Year)

fishfr <- fish %>%
  filter(fished == "no") %>%
  right_join(all) %>%
  arrange(desc(Taxonomy),desc(Year)) %>%
  group_by(Taxonomy) %>%
  mutate(interp = na.approx(Count_m2, na.rm = FALSE)) %>% #don't try this at home! not sure if it is recommended for fill in NAs
  ungroup() %>%
  select(Year, Taxonomy, interp) %>%
  pivot_wider(names_from = Taxonomy, values_from = interp) %>%
  select(-`Ostracion cubicus`) %>% # have to remove this species bc too many missing yrs
  filter(Year != 2018 & Year != 2006) %>% # years with missing data
  as.data.frame()

```

Now we can calculate portfolio effects for the unfished reef fish:

```{r}
library(ecofolio)
pe_avg_cv(fishfr[,-1],ci = TRUE)
pe_mv(fishfr[,-1],ci = TRUE)
```

And, to see how much lower the variability is in the group of unfished reef fish, compared to what you would expect from a population of the same size given the mean-variance relationship, 

```{r}
plot_mv(fishfr[,-1], show = "linear", ci = TRUE)
```

The distance between the X and the O should be the size of the portfolio effect. Usually the X (observed variance) is below the O (expected variance based on mean-variance relationship) but here it is above. This may be due to synchrony among some of the more abundant reef fish. 

### Synchrony among stocks or species
There are several ways to quantify synchrony, but `ecofolio` provides one:  `synchrony()` takes a matrix of biomass data (in our case, `fishfr`) and returns the Loreau and de Mazancourt synchrony index^[Loreau, M., & de Mazancourt, C. (2008). Species synchrony and its drivers: Neutral and nonneutral community dynamics in fluctuating environments. American Naturalist, 172(2), E48–E66. https://doi.org/10.1086/589746]:

$$\phi = \frac{\sigma_{T}^2}{(\sum_{i}\sigma_i)^2}  $$

where $T$ is the 'total' (the whole metapopulation, or whole reef fish community in the case of our example), and $\sigma_i^2$ are the individual population variances. For this index, 0 represents perfect asynchrony and 1 represents perfect synchrony.

For the unfished reef fishes in this cleaned-up dataset, this value is:
```{r}
synchrony(fishfr[,-1])

```
You can tell from this value that the populations are pretty synchronous, which makes sense given the predicted vs. observed variances in the plot above. If you want to dig deeper into whether certain species are synchronous or asynchronous with each other, you can use more sophisticated time series models like MARSS, which accounts for autocorrelation and can include things like environmental drivers. If you were using a MARSS model, you would estimate the off-diagonals of the process error correlation matrix to estimate the degree of synchrony between species (or population components, if that's what you're looking at). More for a later discussion! 
