---
title: "portfolioexample"
author: "Megsie"
date: "9/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Portfolio effects

This doc accompnaies the timesseries.Rmd presentation. I had some issues with `xaringan` so I moved the second half of the demo to a Markdown file. 

## Example with LTER fish data

### The data

Lily has prepared some LTER fish data, which look like this:
```{r echo=FALSE, eval=TRUE, tidy=FALSE}
fish <- read.csv(here::here("data","final_six_fished_unfished.csv"))
fish %>%
  mutate(fished = recode(fished, 
                         no = "unfished",
                         yes = "fished")) %>%
  ggplot(aes(x = Year, y = Count_m2, colour = Taxonomy)) +
  geom_line() +
  facet_wrap(~fished) +
  theme_classic(base_size = 16)
```

### Portfolio analysis using `ecofolio`

This requires data for every year. In the past I have dealt with this by using a state-space model to estimate states for every year, but here I just did some interpolation thing with `zoo::na.approx()` for expediency.

```{r echo=FALSE}
library(zoo) # for filling NAs
all <- fish %>% 
  filter(fished == "no") %>%
  expand(Taxonomy, Year)

fishfr <- fish %>%
  filter(fished == "no") %>%
  right_join(all) %>%
  arrange(desc(Taxonomy),desc(Year)) %>%
  group_by(Taxonomy) %>%
  mutate(interp = na.approx(Count_m2, na.rm = FALSE)) %>% #don't try this at home! not sure if it is recommended for fill in NAs
  ungroup() %>%
  select(Year, Taxonomy, interp) %>%
  pivot_wider(names_from = Taxonomy, values_from = interp) %>%
  select(-`Ostracion cubicus`) %>% # have to remove this species bc too many missing yrs
  filter(Year != 2018 & Year != 2006) %>% # years with missing data
  as.data.frame()

```

Now we can calculate portfolio effects for the unfished reef fish:

```{r}
library(ecofolio)
pe_avg_cv(fishfr[,-1],ci = TRUE)
pe_mv(fishfr[,-1],ci = TRUE)
```

And, to see how much lower the variability is in the group of unfished reef fish, compared to what you would expect from a population of the same size given the mean-variance relationship, 

```{r}
plot_mv(fishfr[,-1], show = "linear", ci = TRUE)
```

The distance between the X and the O should be the size of the portfolio effect, but usually X is below the expected variance. So I'm not sure what's happening here. 