---
title: "Time series analysis"
subtitle: "⚔<br/>with ecofolio and MARSS"
author: "Megsie Siple"
institute: "Stier Lab @ ucsb"
date: "2016/12/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

???

Image credit: [Wikimedia Commons](https://commons.wikimedia.org/wiki/File:Sharingan_triple.svg)

---
class: center, middle

# xaringan

### /ʃaː.'riŋ.ɡan/

---
class: inverse, center, middle

# Get Started

---

# Before we start

Install the **ecofolio** package from [Github](https://github.com/seananderson/ecofolio):

```{r eval=FALSE, tidy=FALSE}
devtools::install_github("seananderson/ecofolio")
```

--

You are recommended to use the [RStudio IDE](https://www.rstudio.com/products/rstudio/), but you do not have to.

- Create a new R Markdown document from the menu `File -> New File -> R Markdown -> From Template -> Ninja Presentation`;<sup>1</sup>

--

- Click the `Knit` button to compile it;

--

- or use the [RStudio Addin](https://rstudio.github.io/rstudioaddins/)<sup>2</sup> "Infinite Moon Reader" to live preview the slides (every time you update and save the Rmd document, the slides will be automatically reloaded in RStudio Viewer.

.footnote[
[1] 中文用户请看[这份教程](http://slides.yihui.org/xaringan/zh-CN.html)

[2] See [#2](https://github.com/yihui/xaringan/issues/2) if you do not see the template or addin in RStudio.
]

---

# Refresher about time series

There are some time series basics that it helps to know when you're analyzing time series data.

These basics are all included in `stats` so you don't have to download a package. 

--- 


# Random walk

Random walk models look like this:

$$X_t = X_{t-1}+w_t$$

In R code, this would look like the following:
```{r}
set.seed(123)
TT <- 100
xx <- ww <- rnorm(n = TT, mean = 0, sd = 1)
for (t in 2:TT) {# compute values 2 thru TT
    xx[t] <- xx[t - 1] + ww[t]
}
```
---

# First-order autoregressive (AR-1) models

AR-1 models look like this:

$$X_t = c + \rho X_{t-1} +\epsilon_t$$
where $X_t$ is the variable of interest (usually abundance or biomass), $c$ is a constant, $\rho$ is an autocorrelation parameter, and $\epsilon_t$ is white noise ( $\epsilon_t ~ N(0, \sigma^2)$ ).

-- 
You can simulate an AR-1 model in R as follows:


---
